---
title: "class11"
format: pdf
editor: visual
---

# Section 1. Proportion of G/G in a population

Read csv file for MXL data

```{r}
#mexican data in LA
mxl <-  read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)

```

```{r}
table(mxl$Genotype..forward.strand.)
#table summarizes data... shows number of each occurance
```

To find proportion, divide by total number of people in data set

```{r}
round(table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100, 2)
#G|G is quite low...
```

read GBR data and find proportion

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100, 2)
```

This variant that is associated with childhood asthma is more frequent in the GBR population than MXL

## Section 4. Population Scale Analysis

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

Sample Size:

```{r}
table(expr$geno)
```

Median Expression

```{r}
median_A_A <- median(expr$exp[expr$geno == "A/A"])

median_A_G <- median(expr$exp[expr$geno == "A/G"])

median_G_G <- median(expr$exp[expr$geno == "G/G"])

print(median_A_A)
print(median_A_G)
print(median_G_G)
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)
```

```{r}
ggplot(expr) + aes(x=geno, y=exp, fill=geno) +
  geom_boxplot(notch=T)
```

A\|A expression value is higher than G\|G.

```{r}
nrow(read.csv("Galaxy3-[HG00109_1.fastq].fastqsanger"))
```

the rs8067378 SNP does seem to have an effect on expression of ORMDL3. When one A is replaced with G, there is less expression. When both are replaced with G, there is even less expression.
